A comparison of different methods for calculating tangent-stiffness matrices in a massively parallel computational peridynamics code.

### 2014-04-25 jacobian_calc_CMAME.tex
Small edits

\section{Introduction}
%\subsection{Motivation}
\label{sec:intro}

@@

        \caption{Displacement, D, after load-step 12} \label{tab:Verification}   

@@

        \caption{Summed iterations over all load steps} \label{tab:ConvergenceStudy}   

@@

\section{Description of analysis tools and approach} 
%
\subsection{Peridigm and peridynamics}
\label{subsec:PaP}
%
Comparisons of the different tangent-stiffness matrix calculation techniques were carried out through code-development and instrumentation of the computational peridynamics code \emph{Peridigm} \cite{peridigm}.  \emph{Peridigm} is distributed by Sandia National Laboratories as its primary open-source computational peridynamics code. It is a massively parallel simulation code for implicit and explicit multi-physics simulations primarily used for solid mechanics and material failure. \emph{Peridigm} is a C++ code utilizing agile software components from Sandia's \emph{Trilinos} project \cite{trilinos}. It's important to note that there is no specialization of the tangent-stiffness matrix calculation methods described previously to this particular code or more generally to a computational peridynamics approach and the analysis carried out in this study should provide insight into the speed and accuracy of these methods when implemented into other computational mechanics analysis tools as well.  \emph{Peridigm} was chosen primarily because the FD, CD, and AD methods for tangent-stiffness matrix calculation where already implemented in the code, leaving only the CS method for development. Also, because peridynamics is a nonlocal theory, the tangent-stiffness matrices have a much higher bandwidth (i.e. less sparsity) than traditional finite element tangent-stiffness matrix computations and the time required to construct the tangent-stiffness matrix is a majority of the total computation time in any quasi-static or implicit dynamics analysis; therefore, the \emph{Peridigm} development team has an interest in profiling the performance of the tangent-stiffness matrix computation methods.

@@

A series of nine single compute core (i.e.\ serial) test problems were run with the discretization size being refined with every test in the series. The aim of increasing the refinement was to examine differences between the methods in terms of accuracy and speed. This series would comprise the single core runs.  \todo[]{Re: II.16} The specific parameters of these tests can be found in \emph{Peridigm} input {\tt xml} files included in the paper repository, however the parameters are reproduced in Table~\ref{tab:TestParams}. These {\tt xml} files allow users with the appropriately modified version of \emph{Peridigm} to reproduce the results found in this paper.
%
\begin{table*}[!tbp]    
  \scriptsize
  \centering
        \caption{Test Parameters} \label{tab:TestParams}   
       \begin{tabular}{c c c c c c}
         \toprule
         Peri. nodes & Nozero TS elems. & Horizon ($\delta$) & Load rate & Number Load steps & Abs. tolerance \\ 
                     & $\times 10^6$    & \si{\centi\meter}& \si{\centi\meter\per\second} &   &     \\
        \midrule
        1000 & $1.99$             & $30$   & $2$ & $2$ & $\num{10E-8}$ \\
        2000 & $4.25$             & $24$  & $2$ & $2$ & $\num{10E-8}$ \\
        3000 & $7.79$             & $21$  & $2$ & $2$ & $\num{10E-8}$ \\
        4000 & $15.4$             & $19$  & $2$ & $2$ & $\num{10E-8}$ \\
        6000 & $20.0$             & $16$  & $2$ & $2$ & $\num{10E-8}$ \\
        8000 & $31.4$             & $15$  & $2$ & $2$ & $\num{10E-8}$ \\
        12000 & $40.1$            & $13.1$ & $2$ & $2$ & $\num{10E-8}$ \\
        16000 & $83.9$            & $12$  & $2$ & $2$ & $\num{10E-8}$ \\
        32000 & $165.0$           & $9.5$ & $2$ & $2$ & $\num{10E-8}$ \\
        1000000 & $1670$          & $3$  & $2$ & $2$ & $\num{10E-8}$ \\
        1000000 & $1670$          & $3$  & $2$ & $2$ & $\num{10E-8}$ \\
        1000000 & $1670$          & $3$  & $2$ & $2$ & $\num{10E-8}$ \\
        1000000 & $1670$          & $3$  & $2$ & $2$ & $\num{10E-8}$ \\
        \bottomrule
    \end{tabular}
\end{table*}

@@

\section{Results and discussion}
\label{sec:Res}
%
Table~\ref{tab:results} shows the compute core count, number of nonzero tangent-stiffness (TS) matrix elements, load step average calculation time and load step average accuracy. The accuracy measure is given by equation~(\ref{eqn:accuracy}) and the choice of elastic modulii mentioned in Section~\ref{tcs}.
%
\begin{table*}[!tbp]    
  \scriptsize
  \centering
        \caption{Averaged Results, Each Test} \label{tab:results}   
       \begin{tabular}{c c c c c c c c c}
         \toprule
         Cores & TS Elements & \multicolumn{4}{c}{Calculation Time ($\si{\second}$)} & \multicolumn{3}{c}{Accuracy Difference ($\si{\mega\pascal}$)} \\ 
         & $\times 10^6$ & CS & CD & FD & AD & CS & CD & FD \\
        \midrule
        1 & $\num{1.99}$  & 3.5 & 2.7 & 1.7 & 1.6 & $\num{1.92E-10}$ & $\num{1.21E-4}$ & .137 \\
        1 & $\num{4.25}$  & 6.2& 4.9& 3.2& 2.9 & $\num{2.28E-10}$ & $\num{9.94E-4}$ & .148 \\
        1 & $\num{7.79}$  & 11.2& 8.9& 5.7& 5.2 & $\num{2.38E-10}$ & $\num{1.59E-4}$ & .145\\
        1 & $\num{15.4}$  & 26.7& 21.0& 13.4& 12.2 & $\num{2.33E-10}$ & $\num{4.61E-4}$ & .12 \\
        1 & $\num{20.0}$  & 28.1& 22.2& 14.4& 13.1 & $\num{2.76E-10}$ & $\num{1.05E-3}$ & .145 \\
        1 & $\num{31.4}$  & 47.6& 37.7& 24.2& 21.9 & $\num{2.64E-10}$ & $\num{1.65E-3}$ & .133 \\
        1 & $\num{40.1}$  & 55.6& 44.1& 28.4& 25.9 & $\num{3.03E-10}$ & $\num{1.92E-3}$ & .148 \\
        1 & $\num{83.9}$  & 138.9& 109.6& 70.2& 64.0 & $\num{3.63E-10}$ & $\num{1.64E-3}$ & .123 \\
        1 & $\num{165.0}$  & 277.3& 218.1& 139.4& 126.5 & $\num{3.26E-10}$ & $\num{2.18E-3}$ & .128 \\
        32 & $\num{1670}$  & 336.1& 277.7& 200.6& 233.1 & $\num{6.21E-10}$ & $\num{1.52E-2}$ & .176 \\
        64 & $\num{1670}$  & 169.9& 140.7& 102.0& 119.7 & $\num{6.20E-10}$ & $\num{1.50E-2}$ & .177 \\
        96 & $\num{1670}$  & 114.7& 95.0& 69.1 & 79.7 & $\num{6.18E-10}$ & $\num{1.50E-2}$ & .177 \\
        128 & $\num{1670}$  & 86.4& 71.8& 52.4 &58.8 & $\num{6.16E-10}$ & $\num{1.47E-2}$ & .177 \\
        \bottomrule
    \end{tabular}
\end{table*}

@@

The solution accuracy results do follow the same ranking seen in the tangent-stiffness accuracy results; however, again as in the validation problem the results differ only slightly. Speculatively this degree of error may be important to those running probabilistic simulations such that the propagation of error may be a concern. Also it is conceivable that if this error is a fixed quantity, insensitive to material properties or loading conditions, it may have a greater impact on simulations involving very low force densities such as for small displacements or very compliant materials. In this case it could be recommended that forward difference be avoided. 
%
\section{\todo[]{Re: I.5}Conclusions}
\label{sec:Conc}
%
%Let's leave this section out and see what the reviewer says.  The paper is getting pretty long.
%
%\subsection{For further studies of tangent-stiffness accuracy in Newton's method}
%\label{subsec:FurtherStudies}
%The question is raised; what is the significance of tangent-stiffness estimation accuracy to the convergence of Newton's method?  The per-iteration and even convergence studies shown here indicate the effect of tangent-stiffness accuracy for only a single datum, a very accurate tangent-stiffness estimate, while the validation problem of Section~XX indicates the effect of using a grossly inaccurate tangent-stiffness. The results here show that for very small tangent-stiffness estimate inaccuracies, convergence rate is not affected, while for grossly inaccurate tangent-stiffnesss Newton's method fails.  Additional confounding factors include that a tangent-stiffness is only first order approximation of an often curved energy surface and that a non-naive implementation of Newton's method includes a line search phase that helps in choosing an optimal iterate given a tangent-stiffness.  Is it worthwhile then to investigate the effect of using moderately inaccurate tangent-stiffnesss, produced deliberately or otherwise, on the convergence rate of Newton's method?

@@

\section{Acknowledgements}
\label{sec:ack}
This work was funded in part by grants from the United States Air Force Office of Scientific Research grant number W911NF-11-1-0208 and National Energy Technology Laboratory grant number DE-FE0010808. The authors acknowledge the Texas Advanced Computing Center (TACC) at The University of Texas at Austin for providing HPC resources that have contributed to the research results reported within this paper. URL: \url{http://www.tacc.utexas.edu}.
%% References
%%
%% Following citation commands can be used in the body text:
%% Usage of \cite is as follows:
%%   \cite{key}         ==>>  [#]
%%   \cite[chap. 2]{key} ==>> [#, chap. 2]
%%

@@

\section{References}

@@

\appendix
\renewcommand*{\thesection}{\Alph{section}}
%
\section{Complex-Step Example}
\label{sec:appendixA}
%
Using $f(x) = \cos(x)$, we have
%
\[
f (x + i h) = \cos(x + i h),
\]
%
or through trigonometric identities equivalently
%
\[
f(x + i h) = \cos(x) \cosh(h) - i \sin(x) \sinh(h).
\]
%
Applying equation (\ref{eqn:complexFirstDeriv})
\[
    \frac{\partial f(x)}{\partial x} = - \frac{\sin(x) \sinh(h)}{h} + \mathcal{O}(h^2).
\]
Since $h$ is arbitrary we desire it to be as small as possible or
\[
\frac{\partial f(x)}{\partial x} = \lim_{h \to 0} - \frac{\sin(x) \sinh(h)}{h},
\]
using l'H\^opital's rule
\[
\frac{\partial f(x)}{\partial x} = \lim_{h \to 0} - \frac{\sin(x) \cosh(h)}{1}.
\]
As $h \to 0$ then $\cosh(h) \to 1$ thereby recovering the exact derivative
\[
\frac{\partial f(x)}{\partial x} =-\sin(x).
\]

\section{Automatic differentiation example}
\label{sec:appendixB}

>>>>

\section{Introduction.}
%\subsection{Motivation}
\label{sec:intro}

@@

  \caption{Displacement, D, after load-step 12.} \label{tab:Verification}   

@@

  \caption{Summed iterations over all load steps.} \label{tab:ConvergenceStudy}   

@@

\section{Description of analysis tools and approach.} 
%
\subsection{Peridigm and peridynamics.}
\label{subsec:PaP}
%
Comparisons of the different tangent-stiffness matrix calculation techniques were carried out through code-development and instrumentation of the computational peridynamics code \emph{Peridigm} \cite{peridigm}.  \emph{Peridigm} is distributed by Sandia National Laboratories as its primary open-source computational peridynamics code. It is a massively parallel simulation code for implicit and explicit multi-physics simulations primarily used for solid mechanics and material failure. \emph{Peridigm} is a C++ code utilizing agile software components from Sandia's \emph{Trilinos} project \cite{trilinos}. It's important to note that there is no specialization of the tangent-stiffness matrix calculation methods described previously to this particular code or more generally to a computational peridynamics approach and the analysis carried out in this study should provide insight into the speed and accuracy of these methods when implemented into other computational mechanics analysis tools as well.  \emph{Peridigm} was chosen primarily because the FD, CD, and AD methods for tangent-stiffness matrix calculation where already implemented in the code, leaving only the CS method for development. Also, because peridynamics is a nonlocal theory, the tangent-stiffness matrices have a much higher bandwidth (i.e. less sparsity) than traditional finite element tangent-stiffness matrix computations and the time required to construct the tangent-stiffness matrix is a majority of the total computation time in any quasi-static or implicit dynamics analysis; therefore, the \emph{Peridigm} development team has an interest in profiling the performance of the tangent-stiffness matrix computation methods.

@@

A series of nine single compute core (i.e.\ serial) test problems were run with the discretization size being refined with every test in the series. The aim of increasing the refinement was to examine differences between the methods in terms of accuracy and speed. This series would comprise the single core runs.  \todo[]{Re: II.16} The specific parameters of these tests can be found in \emph{Peridigm} input {\tt xml} files included in the paper repository, however the parameters are reproduced in Table~\ref{tab:TestParams}. These {\tt xml} files allow users with the appropriately modified version of \emph{Peridigm} to reproduce the results found in this paper.
%
\begin{table*}[!tbp]    
  \scriptsize
  \centering
  \caption{Test Parameters.} \label{tab:TestParams}   
       \begin{tabular}{c c c c c c}
         \toprule
         Peri. nodes & Nozero TS elems. & Horizon ($\delta$) & Load rate & Number Load steps & Abs. tolerance \\ 
                     & $\times 10^6$    & \si{\centi\meter}& \si{\centi\meter\per\second} &   &     \\
        \midrule
        1000 & $1.99$             & $30$   & $2$ & $2$ & $\num{10E-8}$ \\
        2000 & $4.25$             & $24$  & $2$ & $2$ & $\num{10E-8}$ \\
        3000 & $7.79$             & $21$  & $2$ & $2$ & $\num{10E-8}$ \\
        4000 & $15.4$             & $19$  & $2$ & $2$ & $\num{10E-8}$ \\
        6000 & $20.0$             & $16$  & $2$ & $2$ & $\num{10E-8}$ \\
        8000 & $31.4$             & $15$  & $2$ & $2$ & $\num{10E-8}$ \\
        12000 & $40.1$            & $13.1$ & $2$ & $2$ & $\num{10E-8}$ \\
        16000 & $83.9$            & $12$  & $2$ & $2$ & $\num{10E-8}$ \\
        32000 & $165.0$           & $9.5$ & $2$ & $2$ & $\num{10E-8}$ \\
        1000000 & $1670$          & $3$  & $2$ & $2$ & $\num{10E-8}$ \\
        1000000 & $1670$          & $3$  & $2$ & $2$ & $\num{10E-8}$ \\
        1000000 & $1670$          & $3$  & $2$ & $2$ & $\num{10E-8}$ \\
        1000000 & $1670$          & $3$  & $2$ & $2$ & $\num{10E-8}$ \\
        \bottomrule
    \end{tabular}
\end{table*}

@@

\section{Results and discussion.}
\label{sec:Res}
%
Table~\ref{tab:results} shows the compute core count, number of nonzero tangent-stiffness (TS) matrix elements, load step average calculation time and load step average accuracy. The accuracy measure is given by equation~(\ref{eqn:accuracy}) and the choice of elastic modulii mentioned in Section~\ref{tcs}.
%
\begin{table*}[!tbp]    
  \scriptsize
  \centering
  \caption{Averaged Results, Each Test.} \label{tab:results}   
       \begin{tabular}{c c c c c c c c c}
         \toprule
         Cores & TS Elements & \multicolumn{4}{c}{Calculation Time ($\si{\second}$)} & \multicolumn{3}{c}{Accuracy Difference ($\si{\mega\pascal}$)} \\ 
         & $\times 10^6$ & CS & CD & FD & AD & CS & CD & FD \\
        \midrule
        1 & $\num{1.99}$  & 3.5 & 2.7 & 1.7 & 1.6 & $\num{1.92E-10}$ & $\num{1.21E-4}$ & .137 \\
        1 & $\num{4.25}$  & 6.2& 4.9& 3.2& 2.9 & $\num{2.28E-10}$ & $\num{9.94E-4}$ & .148 \\
        1 & $\num{7.79}$  & 11.2& 8.9& 5.7& 5.2 & $\num{2.38E-10}$ & $\num{1.59E-4}$ & .145\\
        1 & $\num{15.4}$  & 26.7& 21.0& 13.4& 12.2 & $\num{2.33E-10}$ & $\num{4.61E-4}$ & .12 \\
        1 & $\num{20.0}$  & 28.1& 22.2& 14.4& 13.1 & $\num{2.76E-10}$ & $\num{1.05E-3}$ & .145 \\
        1 & $\num{31.4}$  & 47.6& 37.7& 24.2& 21.9 & $\num{2.64E-10}$ & $\num{1.65E-3}$ & .133 \\
        1 & $\num{40.1}$  & 55.6& 44.1& 28.4& 25.9 & $\num{3.03E-10}$ & $\num{1.92E-3}$ & .148 \\
        1 & $\num{83.9}$  & 138.9& 109.6& 70.2& 64.0 & $\num{3.63E-10}$ & $\num{1.64E-3}$ & .123 \\
        1 & $\num{165.0}$  & 277.3& 218.1& 139.4& 126.5 & $\num{3.26E-10}$ & $\num{2.18E-3}$ & .128 \\
        32 & $\num{1670}$  & 336.1& 277.7& 200.6& 233.1 & $\num{6.21E-10}$ & $\num{1.52E-2}$ & .176 \\
        64 & $\num{1670}$  & 169.9& 140.7& 102.0& 119.7 & $\num{6.20E-10}$ & $\num{1.50E-2}$ & .177 \\
        96 & $\num{1670}$  & 114.7& 95.0& 69.1 & 79.7 & $\num{6.18E-10}$ & $\num{1.50E-2}$ & .177 \\
        128 & $\num{1670}$  & 86.4& 71.8& 52.4 &58.8 & $\num{6.16E-10}$ & $\num{1.47E-2}$ & .177 \\
        \bottomrule
    \end{tabular}
\end{table*}

@@

The solution accuracy results do follow the same ranking seen in the tangent-stiffness accuracy results; however, again as in the validation problem the results differ only slightly. Speculatively this degree of error may be important to those running probabilistic simulations such that the propagation of error may be a concern. Also it is conceivable that if this error is a fixed quantity, insensitive to material properties or loading conditions, it may have a greater impact on simulations involving very low force densities such as for small displacements or very compliant materials. In this case it could be recommended that forward difference be avoided. 
%
\section{\todo[]{Re: I.5}Conclusions.}
\label{sec:Conc}
%
%Let's leave this section out and see what the reviewer says.  The paper is getting pretty long.
%
%\subsection{For further studies of tangent-stiffness accuracy in Newton's method}
%\label{subsec:FurtherStudies}
%The question is raised; what is the significance of tangent-stiffness estimation accuracy to the convergence of Newton's method?  The per-iteration and even convergence studies shown here indicate the effect of tangent-stiffness accuracy for only a single datum, a very accurate tangent-stiffness estimate, while the validation problem of Section~XX indicates the effect of using a grossly inaccurate tangent-stiffness. The results here show that for very small tangent-stiffness estimate inaccuracies, convergence rate is not affected, while for grossly inaccurate tangent-stiffnesss Newton's method fails.  Additional confounding factors include that a tangent-stiffness is only first order approximation of an often curved energy surface and that a non-naive implementation of Newton's method includes a line search phase that helps in choosing an optimal iterate given a tangent-stiffness.  Is it worthwhile then to investigate the effect of using moderately inaccurate tangent-stiffnesss, produced deliberately or otherwise, on the convergence rate of Newton's method?

@@

\section{Acknowledgements.}
\label{sec:ack}
This work was funded in part by grants from the United States Air Force Office of Scientific Research grant number W911NF-11-1-0208 and National Energy Technology Laboratory grant number DE-FE0010808. The authors acknowledge the Texas Advanced Computing Center (TACC) at The University of Texas at Austin for providing HPC resources that have contributed to the research results reported within this paper. URL: \url{http://www.tacc.utexas.edu}.
%% References
%%
%% Following citation commands can be used in the body text:
%% Usage of \cite is as follows:
%%   \cite{key}         ==>>  [#]
%%   \cite[chap. 2]{key} ==>> [#, chap. 2]
%%

@@

\section{References.}

@@

\appendix
\renewcommand*{\thesection}{\Alph{section}}
%
\section{Complex-Step Example.}
\label{sec:appendixA}
%
Using $f(x) = \cos(x)$, we have
%
\[
f (x + i h) = \cos(x + i h),
\]
%
or through trigonometric identities equivalently
%
\[
f(x + i h) = \cos(x) \cosh(h) - i \sin(x) \sinh(h).
\]
%
Applying equation (\ref{eqn:complexFirstDeriv})
\[
    \frac{\partial f(x)}{\partial x} = - \frac{\sin(x) \sinh(h)}{h} + \mathcal{O}(h^2).
\]
Since $h$ is arbitrary we desire it to be as small as possible or
\[
\frac{\partial f(x)}{\partial x} = \lim_{h \to 0} - \frac{\sin(x) \sinh(h)}{h},
\]
using l'H\^opital's rule
\[
\frac{\partial f(x)}{\partial x} = \lim_{h \to 0} - \frac{\sin(x) \cosh(h)}{1}.
\]
As $h \to 0$ then $\cosh(h) \to 1$ thereby recovering the exact derivative
\[
\frac{\partial f(x)}{\partial x} =-\sin(x).
\]

\section{Automatic differentiation example.}
\label{sec:appendixB}
